The function-sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-24 crelle-dictionary.tex

In order to extend the dictionary from commutative algebraic groups over finite fields to smooth commutative group schemes $G$ over $\FF_q$, we simply replace the local system $\qcs{E}$ on $G$ with a Weil local system while retaining the extra structure $\mu_\qcs{E}$. 
Set ${\bar G} := G\times_{\Spec{\FF_q}} \Spec{{\bar \FF}_q}$ and let $\Frob{G} : \bG \to \bG$ be the Frobenius automorphism for $\bG$.
In this way we are led to the category $\QC(G)$ of {\it quasicharacter sheaves} on $G$. Objects in $\QC(G)$ are triples $(\gqcs{L}, \mu,\phi)$, where $\gqcs{L}$ is a rank-one local system on ${\bar G}$ and $\phi : \Frob{G}^* \gqcs{L}\to \gqcs{L}$ and $\mu : {\bar m}^* \gqcs{L} \iso \gqcs{L} \boxtimes \gqcs{L}$ are isomorphisms of sheaves satisfying a compatibility condition. Morphisms in $\QC(G)$ are then homomorphisms of Weil sheaves which are compatible with the extra structure.

We first test the category of quasicharacter sheaves by examining its objects on \'etale commutative group schemes $G$ over $\FF_q$; \'etale group schemes form a counterpoint to connected algebraic groups, since the component group of a smooth commutative group scheme is an \'etale group scheme. 
We find that the group homomorphism 
\begin{equation}
T_G: \QCiso{G} \to \Hom(G(\FF_q),\EEx)
\end{equation}
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$, where $\Weil{} \subset \Gal({\bar\FF}_q/\FF_q)$ is the Weil group for $\FF_q$.
One can show, using the Kunneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is cyclic.
The fact that $T_G$ may not be an isomorphism should not be too troubling: if quasicharacter sheaves $(\gqcs{L}, \mu,\phi)$ and $(\gqcs{L}', \mu',\phi')$ determine the same character of $G(\FF_q)$ then $(\gqcs{L},\phi)$ and $(\gqcs{L}',\phi')$ are isomorphic as Weil sheaves.
Here we also see the importance of working with Weil sheaves; without them, we would not be able to geometrize the character of the \'etale group scheme $\ZZ$ 
with image $\ell^{\ZZ}$, for example. %given by $1\mapsto \ell$, for example.

@@

By any measure, there are more quasicharacter sheaves for $T$ than quasicharacters of $T(K)$.  
%\todo{Maybe move this from the introduction into the body of the paper somewhere.}
In this regard, we are reminded of the work of David Vogan \cite{vogan:93a}, in which he finds a geometrization of complete Langlands parameters for $p$-adic groups as equivariant perverse sheaves on Vogan varieties and, in the process, is led to study the representations of all the pure rational forms of the $p$-adic group simultaneously.
A similar phenomenon appears in recent work by Joseph Bernstein in which his geometric Ansatz leads to the study of certain sheaves on the stacky classifying space of the $p$-adic group, resulting in a category which appears to be tied to the representations of all the pure rational forms of the $p$-adic group.
Indeed, Bernstein has suggested to us that our category of quasicharacters for $T$ may be tied to characters of all the pure rational forms of $T$.

@@

We close the introduction with an overview of the structure of this paper.
After setting some basic notation in Section~\ref{sec:notation} 
such as $A^*\ceq \Hom(A,\EEx)$ for any group $A$ and using $G$ for a commutative group scheme over $\FF_q$,
in Section~\ref{sec:category} we define the category of quasicharacter sheaves on $G$.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism $\TrFrob{G} : \QCiso{G} \to G(\FF_q)^*$.
We also define pullbacks of quasicharacter sheaves along morphisms of group schemes.  Applying pullbacks to the component group
sequence for $G$ yields the diagram
 \begin{equation} \label{eq:intro-snake}
  \begin{tikzcd}[row sep=20, column sep=20]
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\FF_q)^* \rar & G(\Fq)^* \rar & G^0(\FF_q)^* \rar & 0,
  \end{tikzcd}
  \end{equation}
which is central to the proof of Theorem~\ref{thm:snake}.  Much of Sections~\ref{sec:etale}, \ref{sec:disc-isog} and \ref{sec:main}
are devoted to proving that the rows of \eqref{eq:intro-snake} are exact and determining the kernels and cokernels of the vertical maps.

@@

\begin{proof}
 Let $b_G : {\bar G} \to G$ be the pullback of $\Spec{\bFq} \to \Spec{\Fq}$ along $G\to \Spec{\Fq}$.
 Suppose $(\qcs{E},\mu_\qcs{E})$ in an object in $\bQC(G)$.
 Then $\qcs{E}$ is an $\ell$-adic constructible sheaf on $G$ and
 $b_G^* \qcs{E}$ comes equipped with an isomorphism
 $\phi_\qcs{E} : \Frob{G}^* b_G^*\qcs{E} \to b_G^* \qcs{E}$.
 The functor $\qcs{E} \mapsto (b_G^* \qcs{E},\phi_\qcs{E})$
 from $\ell$-adic constructible sheaves on $G$ to $\ell$-adic constructible sheaves on $G$
 is full and faithful \citelist{\cite{deligne-katz:SGA7.2}*{Expos\'e XIII} \cite{beilinson-bernstein-deligne:81a}*{Prop. 5.2.1}}.
 This functor preserves local constancy, so takes local systems to local systems.
 The isomorphism $\mu \ceq b_{G\times G}^*\mu_\qcs{E}$ satisfies \ref{QC.2}
 with $b_G^*\qcs{E}$ playing the role of $\gqcs{L}$.
 Moreover, $\phi_\qcs{E}$ is compatible with $\mu_\qcs{E}$ in the sense of \ref{QC.3}.
 This construction defines the functor $B_G : (\qcs{E},\mu_\qcs{E}) \mapsto (b_G^*\qcs{E},b_{G\times G}^*\mu_\qcs{E}, \phi_\qcs{E})$
 and also shows that it is full and faithful.
\end{proof}

@@

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf $\qcs{L}$ on $G$ descends to $G$; let $\qcs{E}$ be an object in $\bQC(G)$ for which $B_G(\qcs{E}) \iso \qcs{L}$.
 Since the functor $B_G : \bQC(G) \to \QC(G)$ is full and faithful, $\Aut(\qcs{L}) = \Aut(\qcs{E})$.
From here, Deligne's function-sheaf dictionary for connected commutative algebraic groups over finite fields \citelist{\cite{deligne:SGA4.5}*{Sommes trig.},\cite{laumon:87a}*{1.1.3}} \todo{Fix citelist} gives us all we need, as we briefly recall.
The key observation is that, by \'etale descent, pullback by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ such that the trace of $\Lang^*\qcs{E}$ is $\trFrob{\qcs{E}}^{-1}$ for every $\qcs{E}$ in $\bQC(G)$. 
Using the equivalence given by $\Lang^*$, if $\Lang^*\qcs{E} = V$ with character $\psi$ and isomorphism $V\to V\otimes V$ then 
$\qcs{E} \iso (\Lang_! V)_\psi$, the $\psi$-isotypic component of the push-forward $\Lang_* V$ ($\Lang$ is proper, so $\Lang_* = \Lang_!$) of the constant sheaf $V$ on $G$ equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes\qcs{E}$ determined by $V\to V\otimes V$. 
Moreover, $\Aut(\qcs{E})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{} V \arrow{r}{\rho} & V\arrow{d}{}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{E}) = 1$, completing the proof.
\end{proof}

@@

\begin{theorem}\label{thm:autornaught}
Let $G$ be a smooth commutative group scheme over $\Fq$.
There is a canonical isomorphism
\[
\Aut(\qcs{L}) \iso \Hom(\pi_0(\bG)_{\Weil{}},\EEx).
\]
\end{theorem}
\begin{proof} 
%For the first part of the proposition, argue as in the analogous part of Proposition~\ref{prop:autornaught_etale}.
%For the second, f
Fix $\qcs{L} = (\gqcs{L},\mu,\phi)$ and consider the group homomorphism $\Aut(\qcs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ defined in Proposition~\ref{prop:autornaught_etale}.
This homomorphism is injective because morphisms of sheaves are determined by the linear transformations induced on stalks.  
Note that the image of $\Aut(\qcs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ is contained in $\Hom_\text{cts}(\bG_{\Weil{}},\EEx)$, where $\bG_{\Weil{}}$ is given the topology of the base of the espace \'etal\'e attached to $\gqcs{L}$.
Since $\ell$ is invertible in $\Fq$, it follows that $\Hom_\text{cts}(\bG_{\Weil{}},\EEx) = \Hom(\pi_0(\bG_{\Weil{}}),\EEx)$. 
Note that $\pi_0(\bG_{\Weil{}})=\pi_0(\bG)_{\Weil{}}$. 
To see that $\Aut(\qcs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ is surjective, begin with $\theta\in\Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ and, for each $[x] \in \pi_0(\bG)_{\Weil{}}$ define $\bar\rho^y: \gqcs{L}^y \to \gqcs{L}^y$ by scalar multiplication by $\theta([x])\in \EEx$ for each $y\in [x]$.
This defines an isomorphism $\bar\rho : \gqcs{L}\to \gqcs{L}$ of local systems on $\bG$ compatible with $\mu$ and $\phi$, and thus an isomorphism $\rho :\qcs{L}\to \qcs{L}$ with maps to $\theta$ under $\Aut(\qcs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$.
\end{proof}

@@

Let $K$, $R$ and $R_n$ be as above.
We will need the Greenberg functor, which provides
a functor from the category of schemes 
over $R_n$, for each $n\in \NN$, to the category of schemes over $\Fq$; \todo{font for categories, subscripts on homsets}
See \cite{bertrapelle-gonzales:Greenberg} for the definition and fundamental properties of the Greenberg functor as we use it; other important references include
\citelist{
\cite{greenberg:61}, \cite{greenberg:63a},
\cite{demazure-gabriel:GroupesAlgebriques}*{V, \S 4, no. 1},
\cite{bosch-lutkebohmert-reynaud:NeronModels}*{Ch. 9, \S 6},
\cite{nicaise-sebag:motivicSerre}*{\S 2.2}
}.
%
For any non-negative integer $n$ we will write
\[
\Gr^R_n : \Sch{R} \to \Sch{\Fq}
\]
for the functor produced by composing pullback along $\Spec{R_n} \to \Spec{R}$ with the Greenberg functor. 
This functor respects open immersions, closed immersions, \'etale morphisms, smooth morphisms and geometric components.  Moreover, there is a canonical isomorphism
\[
\Gr^R_n(X)(\Fq) \iso X(R_n)
\]
for any scheme $X$ over $R$.

>>>>

In order to extend the dictionary from commutative algebraic groups over finite fields to smooth commutative group schemes $G$ over $\FF_q$, we simply replace the local system $\qcs{E}$ on $G$ with a Weil local system while retaining the extra structure $\mu_\qcs{E}$. 
In this way we are led to the category $\QC(G)$ of {\it quasicharacter sheaves} on $G$. Objects in $\QC(G)$ are triples $(\gqcs{L}, \mu,\phi)$, where $\gqcs{L}$ is a rank-one local system on ${\bar G} := G\times_{\Spec{\FF_q}} \Spec{{\bar \FF}_q}$ and $\phi : \Frob{G}^* \gqcs{L}\to \gqcs{L}$ and $\mu : {\bar m}^* \gqcs{L} \iso \gqcs{L} \boxtimes \gqcs{L}$ are isomorphisms of sheaves satisfying a certain compatibility condition.  Morphisms in $\QC(G)$ are then homomorphisms of Weil sheaves which are compatible with the extra structure.

We first test the category of quasicharacter sheaves by examining its objects on \'etale commutative group schemes $G$ over $\FF_q$; \'etale group schemes form a counterpoint to connected algebraic groups, since the component group of a smooth commutative group scheme is an \'etale group scheme. 
We find that the group homomorphism 
\begin{equation}
T_G: \QCiso{G} \to \Hom(G(\FF_q),\EEx)
\end{equation}
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$, where $\Weil{} \subset \Gal({\bar\FF}_q/\FF_q)$ is the Weil group for $\FF_q$.
One can show, using the Kunneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is cyclic.
The fact that $T_G$ may not be an isomorphism should not be too troubling: if quasicharacter sheaves $(\gqcs{L}, \mu,\phi)$ and $(\gqcs{L}', \mu',\phi')$ determine the same character of $G(\FF_q)$ then $(\gqcs{L},\phi)$ and $(\gqcs{L}',\phi')$ are isomorphic as Weil sheaves.
Here we also see the importance of working with Weil sheaves; without them, we would not be able to geometrize any character of the \'etale group scheme $\ZZ$ 
with non-compact image, such as $\ell^{\ZZ}$.

@@

By any measure, there are more quasicharacter sheaves for $T$ than quasicharacters of $T(K)$.  
%\todo{Maybe move this from the introduction into the body of the paper somewhere.}
In this regard, we are reminded of the work of David Vogan \cite{vogan:93a}, in which he finds a geometrization of complete Langlands parameters for $p$-adic groups 
%as equivariant perverse sheaves on Vogan varieties 
and, in the process, is led to study the representations of all the pure rational forms of the $p$-adic group simultaneously.
A similar phenomenon appears in recent work by Joseph Bernstein in which his geometric Ansatz leads to the study of certain sheaves on the stacky classifying space of the $p$-adic group, resulting in a category which appears to be tied to the representations of all the pure rational forms of the $p$-adic group \cite{bernstein:vogan_conference}.
Indeed, Bernstein has suggested to us that our category of quasicharacters for $T$ may be tied to characters of all the pure rational forms of $T$.

@@

We close the introduction with an overview of the structure of this paper.
After setting some basic notation in Section~\ref{sec:notation} 
such as $A^*\ceq \Hom(A,\EEx)$ for any abelian group $A$ and using $G$ for a smooth commutative group scheme over $\FF_q$,
in Section~\ref{sec:category} we define the category of quasicharacter sheaves on $G$.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism $\TrFrob{G} : \QCiso{G} \to G(\FF_q)^*$.
We also define pullbacks of quasicharacter sheaves along morphisms of group schemes.  Applying pullbacks to the component group
sequence for $G$ yields the diagram
 \begin{equation} \label{eq:intro-snake}
  \begin{tikzcd}[row sep=20, column sep=20]
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\FF_q)^* \rar & G(\Fq)^* \rar & G^0(\FF_q)^* \rar & 0,
  \end{tikzcd}
  \end{equation}
which is central to the proof of Theorem~\ref{thm:snake}.  Much of Sections~\ref{sec:etale}, \ref{sec:disc-isog} and \ref{sec:main}
are devoted to proving that the rows of \eqref{eq:intro-snake} are exact and determining the kernels and cokernels of the vertical maps.

@@

\begin{proof}
 Let $b_G : {\bar G} \to G$ be the pullback of $\Spec{\bFq} \to \Spec{\Fq}$ along $G\to \Spec{\Fq}$.
 Suppose $(\qcs{E},\mu_\qcs{E})$ in an object in $\bQC(G)$.
 Then $\qcs{E}$ is an $\ell$-adic constructible sheaf on $G$ and
 $b_G^* \qcs{E}$ comes equipped with an isomorphism
 $\phi_\qcs{E} : \Frob{G}^* b_G^*\qcs{E} \to b_G^* \qcs{E}$.
 The functor $\qcs{E} \mapsto (b_G^* \qcs{E},\phi_\qcs{E})$
 from $\ell$-adic constructible sheaves on $G$ to $\ell$-adic constructible sheaves on $G$
 is full and faithful; see \cite{deligne-katz:SGA7.2}*{Expos\'e XIII} or \cite{beilinson-bernstein-deligne:81a}*{Prop. 5.2.1}.
 This functor preserves local constancy, so takes local systems to local systems.
 The isomorphism $\mu \ceq b_{G\times G}^*\mu_\qcs{E}$ satisfies \ref{QC.2}
 with $b_G^*\qcs{E}$ playing the role of $\gqcs{L}$.
 Moreover, $\phi_\qcs{E}$ is compatible with $\mu_\qcs{E}$ in the sense of \ref{QC.3}.
 This construction defines the functor $B_G : (\qcs{E},\mu_\qcs{E}) \mapsto (b_G^*\qcs{E},b_{G\times G}^*\mu_\qcs{E}, \phi_\qcs{E})$
 and also shows that it is full and faithful.
\end{proof}

@@

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf $\qcs{L}$ on $G$ descends to $G$; let $\qcs{E}$ be an object in $\bQC(G)$ for which $B_G(\qcs{E}) \iso \qcs{L}$.
 Since the functor $B_G : \bQC(G) \to \QC(G)$ is full and faithful, $\Aut(\qcs{L}) = \Aut(\qcs{E})$.
From here, Deligne's function-sheaf dictionary for connected commutative algebraic groups over finite fields, as in \cite{deligne:SGA4.5}*{Sommes trig.} or \cite{laumon:87a}*{1.1.3}, gives us all we need, as we briefly recall. \todo{Fix citelist} %\citelist{\cite{deligne:SGA4.5}*{Sommes trig.},\cite{laumon:87a}*{1.1.3}}
The key observation is that, by \'etale descent, pullback by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ such that the trace of $\Lang^*\qcs{E}$ is $\trFrob{\qcs{E}}^{-1}$ for every $\qcs{E}$ in $\bQC(G)$. 
Using the equivalence given by $\Lang^*$, if $\Lang^*\qcs{E} = V$ with character $\psi$ and isomorphism $V\to V\otimes V$ then 
$\qcs{E} \iso (\Lang_! V)_\psi$, the $\psi$-isotypic component of the push-forward $\Lang_* V$ ($\Lang$ is proper, so $\Lang_* = \Lang_!$) of the constant sheaf $V$ on $G$ equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes\qcs{E}$ determined by $V\to V\otimes V$. 
Moreover, $\Aut(\qcs{E})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{} V \arrow{r}{\rho} & V\arrow{d}{}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{E}) = 1$, completing the proof.
\end{proof}

@@

\begin{theorem}\label{thm:autornaught}
Let $G$ be a smooth commutative group scheme over $\Fq$.
There is a canonical isomorphism
\[
\Aut(\qcs{L}) \iso \Hom(\pi_0(\bG)_{\Weil{}},\EEx).
\]
\end{theorem}
\begin{proof} 
%For the first part of the proposition, argue as in the analogous part of Proposition~\ref{prop:autornaught_etale}.
%For the second, f
Fix $\qcs{L} = (\gqcs{L},\mu,\phi)$ and consider the group homomorphism $\Aut(\qcs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ defined in Proposition~\ref{prop:autornaught_etale}.
This homomorphism is injective because morphisms of sheaves are determined by the linear transformations induced on stalks.  
Note that homomorphisms in the image of $\Aut(\qcs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ are continuous where $\bG_{\Weil{}}$ is given the topology of the base of the espace \'etal\'e attached to $\gqcs{L}$.
Since $\ell$ is invertible in $\Fq$, it now follows that the image of $\Aut(\qcs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ is contained in $\Hom(\pi_0(\bG_{\Weil{}}),\EEx)$. 
We also have $\pi_0(\bG_{\Weil{}})=\pi_0(\bG)_{\Weil{}}$. 
To see that $\Aut(\qcs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ is surjective, begin with $\theta\in\Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ and, for each $[x] \in \pi_0(\bG)_{\Weil{}}$ define $\bar\rho^y: \gqcs{L}^y \to \gqcs{L}^y$ by scalar multiplication by $\theta([x])\in \EEx$ for each $y\in [x]$.
This defines an isomorphism $\bar\rho : \gqcs{L}\to \gqcs{L}$ of local systems on $\bG$ compatible with $\mu$ and $\phi$, and thus an isomorphism $\rho :\qcs{L}\to \qcs{L}$ with maps to $\theta$ under $\Aut(\qcs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$.
\end{proof}

@@

Let $K$, $R$ and $R_n$ be as above.
We will need the Greenberg functor, which provides
a functor from the category of schemes 
over $R_n$, for each $n\in \NN$, to the category of schemes over $\Fq$; \todo{font for categories, subscripts on homsets}
See \cite{bertrapelle-gonzales:Greenberg} for the definition and fundamental properties of the Greenberg functor as we use it; other important references include
\cite{greenberg:61}, \cite{greenberg:63a},
\cite{demazure-gabriel:GroupesAlgebriques}*{V, \S 4, no. 1},
\cite{bosch-lutkebohmert-reynaud:NeronModels}*{Ch. 9, \S 6},

\cite{nicaise-sebag:motivicSerre}*{\S 2.2}. %citelist
%
For any non-negative integer $n$ we will write
\[
\Gr^R_n : \Sch{R} \to \Sch{\Fq}
\]
for the functor produced by composing pullback along $\Spec{R_n} \to \Spec{R}$ with the Greenberg functor. 
This functor respects open immersions, closed immersions, \'etale morphisms, smooth morphisms and geometric components.  Moreover, there is a canonical isomorphism
\[
\Gr^R_n(X)(\Fq) \iso X(R_n)
\]
for any scheme $X$ over $R$.
